---
title: "Cyclistic Case Study"
output: 
  rmdformats::downcute:
    self_contained: true
date: "2022-10-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(knitr)
library(rmdformats)
## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
               prompt=FALSE,
               comment=NA,
               message=FALSE,
               warning=FALSE, 
               results="hide")
opts_knit$set(width=75)
```
## Step 1: Loading Data

### Install and load the packages for R. Then import all of the month's files for the past year.

```{r Load data, echo=TRUE, message=FALSE, warning=FALSE, results= 'hide'}

#install.packages('plyr', repos = "http://cran.us.r-project.org")
#install.packages("contrib.url")
#options(repos = list(CRAN="http://cran.rstudio.com/"))


#install.packages("tidyverse", repos = "http://cran.us.r-project.org")
#install.packages("lubridate", repos = "http://cran.us.r-project.org")
#install.packages("ggplot2", repos = "http://cran.us.r-project.org")

library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data

getwd() #displays your working directory
setwd("~/Desktop/Google Course Bike Share Case Study") #sets your working directory to simplify calls to data


Oct21 <- read_csv("202110-divvy-tripdata.csv") #Upload data sets
Nov21 <- read_csv("202111-divvy-tripdata.csv")
Dec21 <- read_csv("202112-divvy-tripdata.csv")
Jan22 <- read_csv("202201-divvy-tripdata.csv")
Feb22 <- read_csv("202202-divvy-tripdata.csv")
Mar22 <- read_csv("202203-divvy-tripdata.csv")
Apr22 <- read_csv("202204-divvy-tripdata.csv")
May22 <- read_csv("202205-divvy-tripdata.csv")
Jun22 <- read_csv("202206-divvy-tripdata.csv")
Jul22 <- read_csv("202207-divvy-tripdata.csv")
Aug22 <- read_csv("202208-divvy-tripdata.csv")
Sep22 <- read_csv("202209-divvy-publictripdata.csv")

```

## Step 2: Combine data

### Compare column names of each file. I hid the results to limit clutter of the report's output.

```{r Combine data, echo=TRUE, message=FALSE, warning=FALSE, results = 'hide'}
colnames(Oct21)
colnames(Nov21)
colnames(Dec21)
colnames(Jan22)
colnames(Feb22)
colnames(Mar22)
colnames(Apr22)
colnames(May22)
colnames(Jun22)
colnames(Jul22)
colnames(Aug22)
colnames(Sep22)
```

### Inspect the data frame

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
str(Oct21)
```

### Stack individual month's data frames into one big data frame. I commented out the view of the data to limit clutter.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
all_trips <- bind_rows(Oct21,Nov21, Dec21, Jan22, Feb22, Mar22, Apr22, May22, Jun22, Jul22, Aug22, Sep22)

#View(all_trips)

```

## Step 3: Data Cleaning

### Inspect the new table that has been created

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics

```

### Added some additional columns of data, such as day, month, year, that provide additional opportunities to aggregate the data.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")


```

### Add a "ride_length" calculation to all_trips (in seconds)

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

### Inspect the structure of the columns

```{r}
str(all_trips)
```

### Convert "ride_length" to numeric so we can run calculations on the data. First let's check to see if the ride_length's column's type is numeric.

```{r}
is.numeric(all_trips$ride_length)

all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```

### The data frame includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative. Here I find negative ride lengths. I have commented out the view of the data to limit the clutter of the report.

```{r}
all_trips[all_trips$ride_length<0, ]

#View(all_trips[all_trips$ride_length<0, ])

nrow(all_trips[all_trips$ride_length<0, ])

```

### As you can see, there are 108 instances where the ride lengths have been negative seconds. I will delete these instances. I will create a new version of the data frame (v2) since data is being removed. Again, commenting out the view of the whole data frame.

```{r}
all_trips_v2 <- all_trips[!(all_trips$ride_length<0),]

#View(all_trips_v2)

```

## Step 4: Analysis

### Descriptive analysis on ride_length (all figures in seconds). I hid the results of the following code chunk.

```{r echo=TRUE, results = 'hide'}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride

```

### The summary function is a more compact way of getting the same information as above.

```{r}
summary(all_trips_v2$ride_length)
```

### Compare members and casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```

### See the average ride time by each day for members vs casual users

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

```

### Notice that the days of the week are out of order. Here, I sort them.

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

### Now, let's run it again

```{r}
x <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

head(x)

```

### Analyze ridership data by type and weekday. Here I find the average ride length of casual riders and members per day.

```{r message=FALSE, warning=FALSE, paged.print=FALSE, results= 'hide'}
y <- all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n(),			#calculates the number of rides and average duration 
  average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)	    #sorts
```

```{r}
head(y)
```

## Step 5: Visualizations

### Visualization for the number of rides by rider type

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```

### Visualization for average duration of rides by rider type

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

## Step 6: Save and export

```{r}
write.csv(x, file = "/Users/Devanshi/Desktop/avg_ride_length.csv")
write.csv(y, file = "/Users/Devanshi/Desktop/ride_length_summary_byday_member_casual.csv")

```
